---
title: "Joining Data in R - Casestudy"
output: html_notebook
---


## This is the worktrough with the **Lahman's Baseball Database** from the DataCammp Course

```{r}
install.packages("Lahman")
require(Lahman)
require(dplyr)
require(dtplyr)
require(purrr)
```

# Data Structure: Inspection and making sense
```{r}
# Setting up the dataset
data("LahmanData")

# Storing views on the dataset
data_nobs <- arrange(structure(data), desc(nobs))
data_vars <- arrange(structure(data), desc(nvar))

# Inspecting the datatables individually
data("LahmanData")
Lahman
```
First we check if there is a common `strong-ID` to connect all the individual data.frames
```{r}
require(purrr)

# Use the reduce function to work on all data.frames in a vectorized fashion. We apply the `intersect()` function
reduce(data, intersect)
```
We see: There is no common ID spanning all data.frames
But there are individual connecting IDs binding the datasets to each other. We can use this to automatically bind all information into a single datatable to perform the analysis

```{r}
data %>% 
        bind_rows(.id="dataframe") %>% 
        group_by(var) %>% 
        summarise(n = n()) %>% 
        filter(n>1) %>% 
        arrange(desc(n))
```
Looking up in which of the tables, the `playerID` variable is present
```{r}
data %>%
        bind_rows(.id="dataframe") %>% 
        filter(var=="playerID") %>% 
        `$`(dataframe)                  # This gets a column of the subset directly

# Exemplifying the use of the `$` in a pipe
data(iris)
iris %>% select(starts_with("Sepal")) %>% 
        filter(Sepal.Width > quantile(Sepal.Width, 0.5)) %>% 
        `$`(Sepal.Width)
```

## Looking at the Salaries

When interested in analyzing a persons salarie for over- or underpayment, we would have to first inspect the dataset for selection bias and skewness. 

We will utilize the following `dplyr` functions:

- `distinct()`                    Returns table with duplicates removed         
- `count(vars = value)`          Counts the number of occurences. Given the vars argument, it returns the count for matches

```{r, eval=FALSE, include=FALSE}
count(iris, vars=Species)

# count(names, vars=band) is equivalent to...
iris %>% group_by(Species) %>% 
        summarise(n())


```

### Subsetting on unique entries
```{r}
# Looking for the list of all players, and their associated IDs
players <- Master %>% 
                select(playerID, nameFirst, nameLast) %>% 
                distinct()

players
```

We see that we have data for 18.589 individual players in this dataset

Lets check if we have salary information for all players, or if we are fighting a bias in dataselection...maybe...
```{r}
players %>% anti_join(Salaries) %>% 
        count()
```
Well, we are missing Salary data for 13768 players in the dataset. 
```{r}
missing_salaries <- count(anti_join(players, Salaries)) / count(players)
sprintf("This amounts to %s percent of our data missing Salaries.",round(missing_salaries,3))
```

Now that we know there is a huge hole in the salary data, let's see if we can explain it. Is it possible that these players somehow did not play (and hence did not earn a salary)?
```{r}
# Using anti_join. These players are neither in the Salaries nor in the Appearance Dataset
forgotten_players <- players %>% 
                anti_join(Salaries, by="playerID") %>% 
                anti_join(Appearances, by="playerID") %>% 
                count()

# Using semi_join. These players are not in the Salaries, but they did take part in games
unpaid_players <- players %>% 
        anti_join(Salaries, by="playerID") %>% 
        semi_join(Appearances, by="playerID") %>% 
        count()

message(sprintf("So we have %s unpaid players who never played, and %s players that went unpaid for their games",
                forgotten_players,
                unpaid_players))
```

Maybe the unpaid players did not continue playing and as such did not receive a full salary to be recorded.
```{r}
players %>% 
        anti_join(Salaries, by="playerID") %>% 
        left_join(Appearances, by="playerID") %>% 
        group_by(playerID) %>% 
        summarize("total_games" = sum(G_all, na.rm=TRUE)) %>% 
        arrange(desc(total_games))


```

Many of the players appeared in thousands of games, so our theory is wrong.Lets check if they acutally played in the games they appeared in. We can check this looking at the `at-bat` variable.
```{r}
players %>% 
        anti_join(Salaries, by="playerID") %>% 
        left_join(Batting, by="playerID") %>% 
        group_by(playerID) %>% 
        summarize("total_games"=sum(AB, na.rm=TRUE)) %>% 
        arrange(desc(total_games))


```

Alright, so all these players participated in many games actively, and are well known and high payed professionals. 
We are definatly dealing with missing data in this dataset.

## Introducing the hall of fame

Players are selected to become Members of the Baseball Hall of Fame based on recommendations from their peers. 
There are many different theories about what distinguishes a Hall of Fame player from his peers. 

We will now have a closer look at this:

### Looking at the application statistics

As with any hall of fame, more players are nominated for membership than are actually admitted.

#### Getting all nominees
Let's see how many players have been nominated for the hall of fame.
```{r}
# Finding the distinct players in HallOfFame
nominated <-
        HallOfFame %>%
        distinct(playerID)

# Number of nominated players
nominated %>% 
        count()

# Complete the data with the players names, and surnames
nominated_full <- nominated %>% 
        left_join(Master, by="playerID") %>% 
        select(playerID, nameFirst, nameLast)

# Return the full list of nominated players
nominated_full
```

#### Getting all inducted players
```{r}
# Get all inducted players
inducted <-
        HallOfFame %>% 
        filter(inducted=="Y") %>% 
        select(playerID) %>% 
        distinct()

# Count the number of inducted players
inducted %>% 
        count()

# Only store the relevant information about the players
inducted_full <- inducted %>%
        left_join(Master, by="playerID") %>% 
        select(playerID, nameFirst, nameLast)

# Return the complete Dataset
inducted_full
```


```{r}
sprintf("The acceptance ratio for the Hall of Fame is %s percent overall", (as.integer(count(inducted))/as.integer(count(nominated))) %>% 
                round(.,2))
```

### Awards : Investigating possible reasons for induction

Maybe the amount of Awards earned by a player is a decisive factor. We will explore the data to investigate this relationship.

```{r}
# Counting the number of awards per player
nAwards <- count(AwardsPlayers, playerID)

# Returning the mean awards for players inducted in the Hall of Fame
nAwards %>% semi_join(inducted, by="playerID") %>% 
        summarize(avg_n = mean(n, na.rm=TRUE))

# Returning the mean n of awards for players nominated for the Hall of Fame
nAwards %>% semi_join(nominated, by="playerID") %>%
        anti_join(inducted, by="playerID") %>% 
        summarize(avg_n = mean(n, na.rm=TRUE))

```

### Salaries: 

Might provide another way to differentiate the two groups.
```{r}

# Check who was nominated but was not inducted
notInducted <- nominated %>%
        setdiff(inducted, by="playerID")


# Calculate the Salaries for nominated players
Salaries %>% 
        semi_join(notInducted) %>% 
        group_by(playerID) %>% 
        summarize(max_salary=max(salary)) %>% 
        summarize(mean(max_salary, na.rm=TRUE))

# max average salaries for the inducted players
Salaries %>% 
        semi_join(inducted, by="playerID") %>% 
        group_by(playerID) %>% 
        summarize(max_salary = max(salary)) %>% 
        summarize(mean(max_salary, na.rm=TRUE))
```


What is wrong with our data? The Salry difference does not seem to be very explanatory?
One rule of the Hall of Fame is that people can not be nominated till 5 years after their retirement. 

Is this reflected in our data?

#### Last year of appearance : Analyzing the Appearances dataframe
```{r}

```
```{r}
Appearances %>% 
        semi_join(nominated, by="playerID") %>% 
        group_by(playerID) %>% 
        summarize(last_year = max(yearID)) %>% 
        left_join(HallOfFame, by="playerID") %>% 
        filter(last_year >= yearID) %>% 
        mutate(overtime = last_year - yearID) %>% 
        select(playerID, overtime, last_year, yearID) %>% 
        arrange(desc(overtime))

```
