---
title: "Filtering join and set operations"
output: html_notebook
---

Filtering joins and set operations combine information from datasets without adding new variables. Filtering joins filter the observations of one dataset based on whether or not they occur in a second dataset. Set operations use combinations of observations from both datasets to create a new dataset.

```{r}
require(dplyr)
require(tidyr)
require(DSR)
```

```{r}
data("iris")

dat1 <- iris %>% fiter(Species =="setosa")

# breaking down the observations by quantiles
quantile_high <- iris %>% filter(Sepal.Length >= quantile(Sepal.Length,0.75))
quantile_med <- iris %>% filter(Sepal.Length >= quantile(Sepal.Length, 0.5) & Sepal.Length < quantile(Sepal.Length, 0.75))
quantile_low <- iris %>% filter(Sepal.Length < quantile(Sepal.Length, 0.5))

# Investigating the datasets
dim(quantile_high)
dim(quantile_med)
dim(quantile_low)

# Comparing to original dataste
dim(iris)
```

The following ways can be used to join the datasets back together

## Mutating joins

This class of joins takes two or multiple talbes and returns a mutated version (changed amount of columns) as a set function result of the original tables.


### Types of Mutating joins: left_join(), right_join(), inner_join(), full_join()
```{r}

# The "standart". Joines the primary with the foreign, keeping all information from the primary
lef_join(primary, foreign, by=c("arg1", "arg2", "etc..."))

# The "reverse standard". Joines the foreign with the primary keeping all information from the primary. Joining to the right
right_join(primary, foreign, by=c("arg", "arg", "arg"))

# The "most exclusive". Only keeps information contained in all joined tables. Drops everything else
inner_join(prim_table, foreign_table, by=c("firstArgument", "SecondArgument"))

# The "most inclusive". Keeps all information, removing duplicates. Filling with "NA"
full_join(primary, foreign, by=c("ar", "arg", "arg","etc.."))
```

## Filtering Joins: semi_join(), anti_join()



### Semi joins: semi_join(), anti_join()

This is a filtering joint, where you can match entries in a dataframe against another one to see which entries will result from a mutating join operation. This operation is also a very handy shorthand procedure to avoid matching long strings of arguments.

```{r}
# Creating a randomn set of IQ values
IQ <- as.data.frame(matrix(rnorm(100, 100,15), 10,10))
names(IQ) <- sprintf("Test%s", 1:10)

hist(as.matrix(IQ))
boxplot(as.matrix(IQ))

# Applying statistical calculations to the columns
lapply(IQ, sd)
sapply(IQ, mean)
sapply(IQ, sd)
sapply(IQ, var) 

IQ %M.7%

```

### anti_join
See which rows will not have a match before joining a dataframe
This can be usefull as a debugging tool for joins...
```{r}
# Returns all entries that are only in the primary table
anti_join(table1, table2, by=c("first", "second", "third", "etc..."))

# Retains only the entries that have a representation in both tables
semi_join()
```


## `Set functions` in R : Combining identical datasets into a new combined dataset

When we have multiple dataframes about the same information, we want to combine them into a complete dataset, only retaining the information for each unique entry. 

This can be done with the Set operations:

- `union()`:            Returns all observations of both datasets, whereby duplicates are removed
- `intersect()`:        Returns only the entries that are represented in both datasets
- `setdiff()`:          Returns only the entries that are represented in the first but not the second

To use them: We pass both datasets 

`union(table1, table2)`

### Exercise to combine their behaviour

There is no funtion to compute an XOR with sets. This way they can be combined to figure this out.
```{r}
# Select songs from live and greatest_hits
live_songs <- live %>% select(song)
greatest_songs <- greatest_hits %>% select(song)

# Return the songs that only exist in one dataset
set_live <- live_songs %>%
setdiff(greatest_songs)

set_greatest <- greatest_songs %>%
setdiff(live_songs)

set_live %>% union(set_greatest)
```


## Comparing datasets

To compare  long datasets we have built in functions in the dplyr package. They return True if the exact same entries are present, in any order.

```{r}
# Returns true only looking at the data, without ordering
setequal(table1, table2)

# This function only returns TRUE if both entries AND ordering are exactly the same
identical(talbe1, talbe2)
```



