---
title: "Working TimeData in R"
output: html_notebook
---
```{r}
require(xts)
require(dplyr)
require(readr)
```



```{r}
data(sunspots)
data()
sunspots_xts <- as.xts(sunspots)

# Investigating the difference
sunspots_xts

# Extracting the core data
core <- coredata(sunspots_xts)
class(core)
core

# Extracting the stimestamps
stamps <-  index(sunspots_xts)
class(stamps)
stamps

# Head and Tail functions to inspect dataset
first(sunspots_xts, "2 months")
last(sunspots_xts, "year")

# Basic subsetting in xts
sunspots_xts["198301/05"]

sunspots_xts[seq(sunspots_xts, "19830101", steps=5)]
```

```{r}
## S3 method for class
'
xts
'
as.xts(x,...,.RECLASS=FALSE)
## S3 method for class
'
timeSeries
'
as.xts(x, dateFormat="POSIXct", FinCenter, recordIDs,
title, documentation, ..., .RECLASS=FALSE)
## S3 method for class
'
its
'
as.xts(x,...,.RECLASS=FALSE)
## S3 method for class
'
zoo
'
as.xts(x, order.by=index(x), frequency=NULL, ..., .RECLASS=FALSE)
## S3 method for class
'
ts
'
as.xts(x, dateFormat,...,.RECLASS=FALSE)
## S3 method for class
'
data.frame
'
as.xts(x, order.by, dateFormat="POSIXct",
frequency=NULL, ...,.RECLASS=FALSE)
## S3 method for class
'
matrix
'
as.xts(x, order.by, dateFormat="POSIXct",
frequency=NULL, ..., .RECLASS=FALSE)
```


### Transforming DateTime objects
```{r}

# Getting start and endpoints for a given daterange
.parseISO8601("2000")

#  start of 2000, end of 2001
.parseISO8601("2000/2001")

# May1, 2000 to end of Feb 2001
.parseISO8601("2000-05/2001-02")

# Jan1, 2000 to Feb 29, 2000 || Truncated shorthand notation!!
.parseISO8601("200001/02")

# 8:30 am to 15:00
.parseISO8601("T08:30/T15:00")

```

### Aligning Time : `align.time()`
lign seconds, minutes, and hours to beginning of next period.
```{r}
# Generate Timeset in second range
index <-  Sys.time() + seq(1,60*60*24*150, 10)
values <- rnorm(length(index), 20000, 2500)
x <- as.xts(values, index)
hist(x)

# change to 10 sec intervals
head(align.time(x, 5), 10)
hist(x)


# align to whole minute
head(align.time(x, 60),10)

# alig to 2 minutes (always have to pass numeric values!!!)
head(align.time(x, 2*60), 10)

# align to every half hour
head(align.time(x, 30*60), 10)

```

### Apply function over time periods:
```{r}
apply.daily()
apply.weekly()
apply.monthly()
apply.quarterly()
apply.yearly()

```


### Lags and Differences of xts Objects: `diff.xts()`

```{r}

## S3 method for class
lag(x, k = 1, na.pad = TRUE, ...)

# xts
diff(x, lag = 1, differences = 1, arithmetic = TRUE, log = FALSE, na.pad = TRUE, ...)

# Example Code
x <- xts(1:10, Sys.Date()+1:10)
lag(x)    # currently using xts-style positive k
lag(x, k=2)
lag(x, k=-1, na.pad=FALSE) # matches lag.zoo(x, k=1)

# 
round(diff(x),2)

# 
diff(x, lag=1)

# 
diff(x, diff=2)

# 
diff(diff(x))
x[1:10]
last(x, "2 hours")
```

Arguments to function call:

- `k` :                   period to lag over
- `differences` :         order of differencing
- `arithmetic` :          should arithmetic or geometric differences be used?


### endpoints: Locate Endpoints by Time

Extract index values of a given xts object corresponding to the
last observations given a period specified by on.

```{r}
x
endpoints(x)

endpoints(x, "weeks")
first(sunpoints_xts, 10)
```

#### first(), last() : Subsetting Timeobjects on time
```{r}
first(1:100)
last(1:100)
LakeHuron
data(LakeHuron)
first(LakeHuron,10)
last(LakeHuron)
x <- xts(1:100, Sys.Date()+1:100)
x
first(x, 10)

# 
first(x, "1 days")

# 
first(x, "4 days")
```


```{r}
# use the proper combination of split, lapply and rbind
temps_1 <- do.call(rbind, lapply(split(temps, "weeks"), function(w) last(w, n = "1 day")))

# create last_day_of_weeks using endpoints()
last_day_of_weeks <- endpoints(temps, "weeks")

# subset temps using last_day_of_weeks 
temps_2 <- temps[last_day_of_weeks]
```

### Time series aggregation

Usefol to convert a univariate series to range bars

```{r}
to.period(x, 
          period="months",
          k = 1,
          indexAt, 
          name = NULL, 
          OHLC = NULL)

# Aggregate data over time periods set as endpoints
to.period(edhec["1997/2001", 1], "years", name="EDHEC")

# Set the Aggregate to be calculated at the beginning of the period
to.period(edhec["1997/2001", 1], "years", name="EDHEC", indexAt = "firstof")
```

### Downsampling data with Dataranges

- force a univariate seris in `to.period()`
```{r}
# OHLC = FALSE
to.period(object[, j], period = "years",
          name="NAME", OHLC = FALSE)

# Apply method
to.period(edhec[,1], period="years", name="EDHEC", OHLC = FALSE)
```

- extract the period values directly
```{r}
# Extract directly
object[endpoints(object, "years"), j]

# Application of this procedure
edhec[endpoints(edhec, "years"), 1]
```


### Rolling function : Apply windwing functions to time series data

Discrete rolling windows

- `split` to break up by period
- `lapply` cumulative functions
        - sumsum, cumprod, cummin, cummax

```{r}
# First split the data into chunks by setting f="timeperiod"
edhec.yrs <- split(edhec[,1], f="years")

# Then apply over the resulting list a cummulative function
edhec.yrs <- lapply(edhec.yrs, cumsum)

# Then `rbind` the values back together
edhec.ytd <- do.call(rbind, edhec.yrs)

# 
cbind(edhec.ytd, edhec[,1]["2007-10/2008-03"])
```

**Continous rolling** with the `rollapply()` function in R :

This applies a cummalative function over a fixed window, adding and droping the nex/last observation as the window is rolled over the list passed as data.

- `data` is the xts object
- `width` is the window size to be applied

```{r}
# Syntax for the rollapply function
rollapply(data, width, FUN, ...,
          by = 1, by.column = TRUE, 
          fill = if(na.pad) NA, 
          na.pad = TRUE, partial = TRUE, 
          align = c("right", "center", "left")

```

Application examples:
```{r}
# Applying a rolling mean over the data from 2007/01 - 2007/08 taking one observation prior or after the current value.
rollapply(edhec["200701/08", 1], 3, mean)
```

Calculating cumsum over years:
```{r}
# Split edhec into years
edhec_years <- split(edhec , f = "years")

# Use lapply to calculate the cumsum for each year in edhec_years
edhec_ytd <- lapply(edhec_years, FUN = cumsum)

# Use do.call to rbind the results
edhec_xts <- do.call(rbind, edhec_ytd)
```


## Looking at the internal function of XTS

### What is the index internally?

All Dates are stored as fractional seconds since "1970-01-01 UTC". The index takes this numeric values, to create the data datetime object xts. 


- `tzone` or `indexTZ`:  represents the Timezone attribute
- `tclass` or `indexClass` :  represents the attribute for extraction
- `indexFormat`: optional display pereferences

```{r}
indexClass(Z)

indexTZ(Z)

indexFormat(Z) <- "%b ,%d, %Y"
```

### Working with TimeZones

Timezones describe time offset from UTC(aka Greenwich Mean Time)

- Always set your time zone to avoid surprises!

```{r}
Sys.setenv(TZ = "America/Chicago")
Sys.setenv(TZ = "Europe/Vienna")

# Set two timestamps with different TZ sets 
ts_vienna <- Sys.time()
ts_chicago <- Sys.time()

# compare the timestamps
ts_vienna
ts_chicago
```

```{r}
# Check the index class used
indexClass(temps)

# Check the timezone used
indexTZ(temps)

# Change the format for time display
indexFormat(temps) <- "%d/%m/%Y"
```

Workint timezones
```{r}
# Construct times_xts with tzone set to America/Chicago
times_xts <- xts(1:10, order.by = times, tzone = "America/Chicago")

# Change the time zone of times_xts to Asia/Hong_Kong
tzone(times_xts) <- "Asia/Hong_Kong" 
  
# Extract the current time zone of times_xts
indexTZ(times_xts)
```


## Periods, Periodicity and Timestamps

### Periodicity

- Identifies the regularity of data
- Answers: what type of data is present
- Less usefull for irregular timestamps
- Summary measure for the Index

```{r}
# Check the regularity of the dataset
periodicity(sunspots_xts)

# Check on regularity when changing chunksize
periodicity(to.yearly(sunspots_xts))

# Used as a s
```

Counting periods:

- Estimate number of periods(e.g. days, weeks, etc..)
- Irregular series equals irregular counting
- Counting periods greater than periodicity

```{r}
nseconds(x)
nminutes(x)
nhours(x)
ndays(x)
nweeks(x)
nmonths(x)
nquarters(x)
nyears(x)
```



```{r}
# the see number of seconds since 1970-01-01T00:00
.index(sunspots_xts)

# returns the endpoints for a given range in seconds passed sinces 1970-01-01T00:00
.indexday(sunspots_xts)
.indexweek(sunspots_xts)

```

Removing duplicate entries from the index with `make.unique.index()`
```{r}
# Make z have unique timestamps
z_unique <- make.index.unique(z, eps = 1e-4)
z
z_unique
# Remove duplicate times in z
z_dup <- make.index.unique(z, drop = TRUE)
z_dup
# Round observations in z to the next time
z_round <- align.time(z, n = 60)

```

