---
title: "Advanced DataVisualization with ggplot"
output: html_notebook
---

## Advanced DataVisualization with ggplot - Coordinates and Facets

this talks about the coord_ function family. 

We can : 

- zoom in 

Set basic zoom limits to the coordinate system we want to display
```{r}
# ggplot sets automatic limits to the dataset, according to the data available
ggplot(iris, aes(Sepal.Length, Sepal.Width, col = Species)) + 
  geom_point()

# We can set the data to use uniform coordinates instead
ggplot(iris, aes(Sepal.Length, Sepal.Width, col = Species)) + 
  geom_point() + 
  coord_equal(ylim = c(2, 8), xlim = c(2,8))
  

# First: Set limits without cutting of values
ggplot(iris, aes(Sepal.Length, Sepal.Width, col = Species)) + 
  geom_point() + 
  geom_smooth(method = "loess", se = T, span = 0.9)  + 
  coord_cartesian(xlim = c(5.3,6.5)) 

# 

```

## Creating Pie Charts - Using the Polar coordinate system

We can transform data to a polar, circular represenation using the `coord_polar()` system. 

```{r}
# This creates a dummy variable to fill the proportion of the cyl factors
thin.bar <- ggplot(mtcars, aes(1, fill = factor(cyl))) + 
  geom_bar()

# We turn this 
thin.bar

# Into a pie chart, by plotting it along a polar coordinate system and specify `theta` to be "y"
thin.bar + coord_polar(theta= "y")

```


## Facet Layers

They are a usefull layer to produce small multiples, a concept popularized by Edward Tufte in his 1983 Book "Visualization of Quantitative Information".

The basic idea is, to keep the limits of a plot identical and spread the data over a multitude of parallel plots, broken down according to a factor level. 

```{r}
ggplot(iris, aes(Sepal.Length, Sepal.Width, col = Species)) + 
  geom_point() + 
  facet_grid(. ~ Species)

# We can split based on columns, as above, or along the rows
ggplot(iris, aes(Sepal.Length, Sepal.Width, col = Species)) + 
  geom_point() + 
  facet_grid(Species ~ .)
```

And this can be split both on columns and variables, specifying two categorial variables to compare.
```{r}
str(mtcars)

ggplot(mtcars, aes(wt, mpg, col = hp)) + 
  geom_point() + 
  facet_grid(am ~ cyl)
```



### Plotting multiple variables at once

This exercise shows how to plot too many variables at once, using facets, aestethic matches

Its just plain ugly!!!
```{r}
# Import the brewer palet package
if(!require(RColorBrewer)){install.packages("RColorBrewer"); library(RColorBrewer)}

# Code to create the cyl_am col and myCol vector
mtcars$cyl_am <- paste(mtcars$cyl, mtcars$am, sep = "_")
myCol <- rbind(brewer.pal(9, "Blues")[c(3,6,8)],
               brewer.pal(9, "Reds")[c(3,6,8)])

# Basic scatter plot, add color scale:
ggplot(mtcars, aes(x = wt, y = mpg, col = cyl_am)) +
  geom_point() + 
  scale_color_manual(values = myCol)

  
# Facet according on rows and columns.
ggplot(mtcars, aes(wt, mpg, col = cyl_am)) + 
  geom_point() + 
  scale_color_manual(values = myCol) + 
  facet_grid(gear ~ vs)




# Add more variables
ggplot(mtcars, aes(wt, mpg, col = cyl_am, size = disp)) + 
  geom_point() + 
  scale_color_manual(values = myCol) + 
  facet_grid(gear ~ vs)
```

## Freeing rows or columsn which are not present in subplots made by facetting

When we facet on a variable that does represent only a single category, we end up with lots of duplicates.
We can remove them, as we would with `rm.na`, by specifying the axis to free in the function call to `facet_grid()`

```{r}
# Basic scatter plot
ggplot(mamsleep, aes(time, name, col = sleep)) + 
  geom_point()

# Facet rows accoding to vore
ggplot(mamsleep, aes(time, name, col = sleep)) +
  geom_point() + 
  facet_grid(vore ~.)

# Specify scale and space arguments to free up rows
ggplot(mamsleep, aes(time, name, col = sleep)) + 
  geom_point() +
  facet_grid(vore ~. , scale = "free_y", space = "free_y")
```

