---
title: "R Notebook"
output: html_notebook
---

### Data Visualization in ggplot2

There is exploratory data analysis 


```{r}
library(MASS)
library(ggplot2)
mammals


# Visualizing it
ggplot(mammals, aes(body, brain)) +
  geom_point() + 
  stat_smooth(method="lm", col="red",
              se=FALSE)


# 
```

Creating a explanatory plot
```{r}
library(scales)
ggplot(mammals, aes(body, brain)) +
  annotation_logticks() +
  geom_point(alpha=0.6) +
  coord_fixed(xlim=c(10^-3, 10^4), ylim=c(10^-1, 10^4)) +
  scale_x_log10(expression("Body weight (log"["10"]*"(Kg))"),
                breaks=trans_breaks("log10", function(x) 10^x),
                labels=trans_format("log10", math_format(10^.x))) + 
  scale_y_log10(expression("Brain weight (log"["10"]*"(g))"),
                breaks= trans_breaks("log10", function(x) 10^x),
                labels= trans_format("log10", math_format(10^.x))) + 
  stat_smooth(method="lm", col="red", se = FALSE, size = 1)  + 
  theme_classic()




```


Plotting the classic iris dataset
```{r}
data(iris)
iris

# Displaying the initial dataset
ggplot(data = iris, aes(Sepal.Length, Sepal.Width, colour=Species)) +  
  geom_point() + 
  scale_x_continuous(expression("Lenght (cm)")) + 
  scale_y_continuous(expression("Width(cm)"))

```


# Wrappint the x variable as a factor removes the display of all not existent values 
```{r}
ggplot(mtcars, aes(factor(cyl), mpg)) + 
  geom_point()
```

```{r}
ggplot(mtcars, aes(wt, mpg, col=disp)) + 
  geom_point()
```

```{r}
ggplot(mtcars, aes(wt, mpg, size=disp)) + 
  geom_point()
```

This could only be mapped to a categorical variable
```{r}
ggplot(mtcars, aes(wt, mpg, shape=disp)) + 
  geom_point()
```

```{r}
levels(iris$Species) <-  unique(iris$Species)

ggplot(iris, aes(Sepal.Length, Sepal.Width)) +
  geom_jitter(alpha = 0.6) + 
  facet_grid(. ~ Species) + 
  stat_smooth(method="lm", se=FALSE, colour="red") + 
  scale_y_continuous("Seapl Width (cm)",
                     limits=c(2,5),
                     expand = c(0,0)) + 
  scale_x_continuous("Sepal Length (cm)",
                     limits= c(4,8),
                     expand = c(0,0)) + 
  coord_equal()
  
  
```


Adding the Theme Layer
```{r}
levels(iris$Species) <-  unique(iris$Species)

ggplot(iris, aes(Sepal.Length, Sepal.Width)) +
  geom_jitter(alpha = 0.6) + 
  facet_grid(. ~ Species) + 
  stat_smooth(method="lm", se=FALSE, colour="red") + 
  scale_y_continuous("Seapl Width (cm)",
                     limits=c(2,5),
                     expand = c(0,0)) + 
  scale_x_continuous("Sepal Length (cm)",
                     limits= c(4,8),
                     expand = c(0,0)) + 
  coord_equal() +
  

# Adding the Theme to the layer
  theme(panel.background = element_blank(),
        plot.background = element_blank(), 
        legend.background = element_blank(),
        legend.key = element_blank(),
        strip.background = element_blank(),
        axis.text = element_text(colour = "black"),
        axis.ticks = element_line(colour = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "black"),
        strip.text = element_blank(),
        panel.spacing = unit(1, "lines")
        
        
        
        )
```

```{r}
str(diamonds)


ggplot(diamonds, aes(carat, price), colour=cut) + 
  geom_point() + 
  geom_smooth(se=TRUE, colour="red")
```


```{r}
ggplot(diamonds, aes(carat, price, col=clarity)) + 
  geom_smooth()
```

```{r}
plot1 <- ggplot(diamonds, aes(carat, price, col=clarity))

plot1 + geom_point(alpha=0.2)
```

```{r}


```

### Working with multiple layers on an object
```{r}
p <- ggplot(iris, aes(Sepal.Length, Sepal.Width))
p + geom_point()
p + geom_jitter()

```

## Using the base package / vs the ggplot2 Package for plotting
```{r}

# Drawing a subset plot with linear regression lines for the given dataset


mtcars$cyl <- as.factor(mtcars$cyl)
plot(mtcars$wt, mtcars$mpg, col = mtcars$cyl)

# Draw the lm model as line
carModel <- lm(mpg ~ wt, data=mtcars)
abline(carModel, lty=2)

# Plot the model individually for each subset
lapply(mtcars$cyl, function(x){
  abline(lm(mpg ~ wt, mtcars, subset = (cyl == x)), col = x)
})

# Draw the legend
legend(x = 5, y = 33, legend = levels(mtcars$cyl),
       col = 1:3, pch = 1, bty = "n")
```

## Doing the same with the ggplot package
```{r}
ggplot(mtcars, aes(wt, mpg, col=cyl)) +
  geom_point() + 
  geom_smooth(method="lm", se=FALSE) + 
  geom_smooth(method="lm", se=FALSE, aes(group=1), lty=2)
```

