---
title: "R Notebook"
output: html_notebook
---

## Proper Data Format in ggplot2

```{r}
require(ggplot2)
require(caret)
```
## Plotting multiple variables 


Instead of doing a duplicated point call to overplot a plot with 4 variables...
```{r}
ggplot(iris, aes(Sepal.Length, Sepal.Width)) + 
  geom_point() + 
  geom_point(aes(Petal.Length, Petal.Width, col="red"))
```

Which still does not display a correct legend to the dataset..
We need to keep the standard ggplot call:

- Taking three variables (Length, Width and LeafType)
- and plot them onto three dimensions(x,y,colour)

For this procedure to work, we first need to arrange the data accordingly
```{r}




```

```{r}
require(tidyr)

# Cleaning the dataset
iris.tidy <- iris %>% 
      gather(key, Value, -Species) %>% 
      separate(key, c("Part", "Measure"), "\\.")

# Inspecting the dataset
str(iris.tidy)


# Drawing the plot
ggplot(iris.tidy, aes(Species, Value, col=Part)) + 
  geom_jitter() +
  facet_grid(. ~ Measure)
  

```

Creating different collections for the dataset
```{r}
head(iris)
```


```{r}
data("iris")

iris$Flower <- 1:nrow(iris)

# Creating the dataset
iris.wide <- iris %>% 
      gather(key, value, -Species, -Flower) %>% 
      separate(key, c("Part", "Measure"), "\\.") %>% 
      spread(Measure, value)

# inspecting the dataset
str(iris.wide)

# Plotting Width / Length over Species
ggplot(iris.wide, aes(Length, Width, col=Part)) + 
  geom_jitter() + 
  facet_grid(.~ Species)

# Plotting width/ Length by 
str(iris.wide)
ggplot(iris.wide, aes(Width, Length, shape=Species, col=Species)) + 
  geom_jitter() + 
  facet_grid(.~ Part)
``` 

## Exploring Aesthetics

For ggplot 2 these are attributes. 
These are used what variables are mapped onto what visual astetic qualities

for example :
- colour
- size 
- 


Modifying Asthethics:
```{r}

ggplot(mtcars, aes(factor(cyl), fill=factor(am))) +
  geom_bar(stat="count", position="stack")

```
```{r}
ggplot(mtcars, aes(factor(cyl), fill=factor(am))) + 
  geom_bar(stat="count", position="fill")
```
```{r}
ggplot(mtcars, aes(factor(cyl), fill=factor(am))) +
  geom_bar(stat="count", position="dodge")
```
Setting additional parameters
```{r}
# Setting the values
val = c("#E41A1C", "#377EB8")
lab = c("Manual", "Automatic")

# Drawing the plot
ggplot(mtcars, aes(factor(cyl), fill=factor(am))) +
  geom_bar(stat="count", position="dodge") +
  scale_x_discrete("Cylinders") +
  scale_y_continuous("Number") +
  scale_fill_manual("Transmission", values = val,
                    labels= lab) 
```


## Setting a dummy aesthetic

Jitter introduces a random noise with a set sd of 0.2, this is used to artificially set the dots appart so they do not overplot on the same exact plot location.

```{r}
mtcars$group <- 0

# Plot the data against 0
ggplot(mtcars, aes(mpg, group)) + 
  geom_jitter() + 
  scale_y_continuous(limits = c(-2,2))
```

### Aesthetics best practice

Which one do we choose, but there is clear guidance. 

refers to: Jacques Bertin, William Cleveland.

You want to either explore, or to educate and persuade. This is defined based on the appropriate audience for the plotting.


### Aesthetics - Continous Variables
```{r}
ggplot(iris, aes(Sepal.Length, Sepal.Width, col=Species)) + 
  geom_point()
```

Now mapping the continous "Widt" Element onto colour.
This is not at all practical, and in the best case the reader will interpret the data in a completly different way.


```{r}
ggplot(iris, aes(Species, Sepal.Width, col=Sepal.Length)) + 
  geom_point()
```

Aesthetics include:

- x
- y
- size 
- shape
- fill
- colour
- alpha


Definatly avoid:

- unaligned scales
- adding duplicated visual differences (col&shape)
- 

### Overplotting Errors
```{r}
ggplot(mtcars, aes(wt, mpg, col=cyl)) + 
  geom_point(size=4)


ggplot(mtcars, aes(wt, mpg, col=cyl)) + 
  geom_point(shape=1, size=4)

ggplot(mtcars, aes(wt, mpg, col=cyl)) + 
  geom_point(alpha=0.6, size=4)
```

```{r}
ggplot(diamonds, aes(carat, price, col=clarity)) + 
  geom_point()

ggplot(diamonds, aes(carat, price, col=clarity)) + 
  geom_point(alpha=0.5)

ggplot(diamonds, aes(clarity, carat, col=price)) + 
  geom_point(alpha=0.5)

ggplot(diamonds, aes(clarity, carat, col=price)) +
  geom_point(position="jitter", alpha=0.5)
```

### Scatter Plots
Looking at the Geometry layer. 

There are currently 37 different options:

- abline
- area
- bar 
- bin2d 
- blank
- boxplot
- contour
- crossbar
- density
- density2d
- dotplot
- errorbar
- errorbarh
- freqpoly
- hex
- histogram
- hline
- jitter
- line
- linerange
- map
- path
- point
- pointrange
- polygon
- quantile
- raster
- rect
- vline
- ribbon
- rug
- segment
- smooth
- step
- text
- tile
- violin


Lets look at the common plots:

#### Scatter Plot
```{r}
plot1 <- ggplot(iris, aes(Sepal.Length, Sepal.Width)) + 
      geom_point(aes(col=Species))

plot1
```

We can assign the data to use inside a function call to the geom. 

Here we add the data
```{r}
data(iris)
iris.summary <- aggregate(iris[1:4], list(iris$Species), mean)
names(iris.summary)[1] <- "Species"

# Plotting the dataset
plot1 + geom_point(data=iris.summary, shape=15, size=5, aes(col=Species))
```


### Non-inheritance when we switch between differetnt aesthetic categories

Trying to draw a line into the scatterplot, we will get a wrong layout
```{r}
# This results in an erros
# plot1 + geom_vline(data=iris.summary)

# The specific aesthetic for the vline "xintercept" has to be specified
plot1 + geom_vline(data=iris.summary, aes(xintercept= Sepal.Length))

# This can be additionally combined with hline
plot1 + 
  geom_hline(data = iris.summary, aes(yintercept = Sepal.Width)) + 
  geom_vline(data=iris.summary, aes(xintercept = Sepal.Length))
```


Two make this more usefull, we have to colour code the crosshair and visually put it back a little
```{r}
plot1 + 
  geom_hline(data = iris.summary, linetype=2, 
             aes(yintercept = Sepal.Width,
                 col=Species)) + 
  geom_vline(data=iris.summary, 
             aes(xintercept = Sepal.Length, col=Species), 
             linetype=2)
```

Further exercises:
```{r}
ggplot(mtcars, aes(factor(cyl), wt)) + 
  geom_point()

# To conquer overplotting we specify the jitter
ggplot(mtcars, aes(factor(cyl), wt)) + 
  geom_jitter()

# We Define the position to use jitter
posn.j <- position_jitter(widt=0.05)

ggplot(mtcars, aes(factor(cyl), wt)) + 
  geom_point(position=posn.j)
```


### Barplots - Typical Use and Aesthetics
```{r}
ggplot(iris, aes(Sepal.Width, fill = Species)) + 
  geom_histogram(aes(y = ..density..),binwidth = 0.1, alpha=0.7)
```
```{r}
ggplot(iris, aes(Sepal.Width, fill = Species)) + 
  geom_histogram(aes(y = ..density..),binwidth = 0.1, position="dodge")
```

```{r}
ggplot(iris, aes(Sepal.Width, fill = Species)) + 
  geom_histogram(aes(y = ..density..),binwidth = 0.1, position="fill")
```

### Bar-Plots

```{r}
# Looking at the sleeping behaviour based on eating habits
ggplot(msleep, aes(vore)) + geom_bar()

```

```{r}

# Chaning the y aes to density, searches for the density variable in the resulting internal stats table
ggplot(mtcars, aes(mpg)) + 
  geom_histogram(binwidth=1, fill="lightblue", aes(y=..density..))

# Now in custom colour
myBlue <- "#377EB8" 
ggplot(mtcars, aes(mpg)) + 
  geom_histogram(binwidth=1, fill=myBlue, aes(y=..density..))


# This displays a continous density curve
ggplot(mtcars, aes(mpg)) + 
geom_histogram(stat = "density")
```

Using the built in position commands to specifiy further details as arguments to the position call
```{r}
# The standard result uses "stack"
ggplot(mtcars, aes(cyl, fill=factor(am))) + 
  geom_bar()

# When we set to "dodge", we get a unhappy "next to each other"
ggplot(mtcars, aes(cyl, fill=factor(am))) +
  geom_bar(position="dodge")

# Using the "width" setting for the dodge command we can let them overlay
pos_d <- position_dodge(width=0.2)
ggplot(mtcars, aes(cyl, fill=factor(am))) + 
  geom_bar(position=pos_d, alpha=0.6)

```

The "freqpoly" frequency Polynom Layer
```{r}
ggplot(mtcars, aes(mpg, col=cyl)) + 
  geom_freqpoly(binwidth=1)
``` 


### Using set "Color Paletts" with `scale_fill_brewer()`
```{r}
ggplot(mtcars, aes(mpg, fill=factor(am))) + 
  geom_bar(binwidth=1,position="dodge") +
  scale_fill_brewer()
```
```{r}
# Import library
if(!require(RColorBrewer)){install.packages("RColorBrewer"); library(RColorBrewer)}

# Create individual color scales
blues <- brewer.pal(9, "Blues")

blue_range <- colorRampPalette(blues)
  
```

### Work