---
title: "Advanced Joining in R"
output: html_notebook
---
```{r}
require(dplyr)
```

## Advanced Joines in R: Problems and solutions

Join multiple types of data


### Pitfalls in Joining: Common problems in data handling


- missing keys (either values or columns missing)
- duplicate keys (either values or columns duplicated)

### Missing Key values:

Occurs when data is missing in the primary key column of your table. NAs introduced.
The required information for a joint is just not available, if it can not be introduced trough inference from other information
these lines have to be dropped.

```{r, eval=FALSE, include=FALSE}
namesNA %>% 
        filter(!is.na(name)) %>% 
        left_join(plays, by="name")
```

A missing key column occurs, we can still try to look at the rownames, and use the `rownames_to_column(noNames, var="name")` to 
transform them into a column we can use to do the join. 

When we have duplicate keys and we have to join them, we  will rapidly increase the amount rows due to a return of all 
possible key value combinations in the dataset.


```{r}
require(tibble)
rownames_to_column("colName")
```

When joining two tables where the keys are not named right, we sometimes have to set the match ourselves
```{r, eval=FALSE, include=FALSE}
left_join(table1, table2, by=c("name" = "writer", "job"="occupation"))
```

When we have two columns that are named identically in both datasets, but are not a `key column`, we can specify the key column to match.
Automatically, all other columns will be set to be values and will be named repeatedly

```{r}
df1 <- data_frame(
        "name" = c("Mick", "John", "Paul"),
        "plays"=c("Stones", "Beatles", "Beatles")
)
df2 <- data_frame(
        "name"=c("John", "Paul", "Keith"),
        "plays"=c("Guitar", "Bass", "Guitar")
)

# Left join results in
left_join(df1, df2, by="name")
```
The suffix .x and .y are automatically added to indicate which dataset the information originated from.
We can change this standard behaviour by setting the suffic argument
```{r}
tbl_joined <- left_join(df1, df2, by="name", suffix=c("1", "2"))
tbl_joined
```

Then rename the columns any other way
```{r}
tbl_joined <- tbl_joined %>% rename("Band"=plays1, "Instrument"=plays2)
tbl_joined
```

When joining two datasets with mismatched key names, we can set them accordingly by hand.
```{r, eval=FALSE, include=FALSE}
x %>% left_join(y, by=c("x.name"="y.name"))
```

## When joining 3 or more datasets

Joins are defined to work with two datasets. As such, joining multiple tables could be executed in a repetitive fashion, using the pipe operator.

```{r, eval=FALSE, include=FALSE}
x %>% 
        left_join(y) %>% 
        left_join(z) %>% 
        lef_join(etc...)
```



## using the `purrr` package 

also created by Hadley Wickham that uses a strategy of list apply to apply datastructure function to a list of objects.

Its main goal is to help you **efficiently apply R functions to data**. 

Its syntax is straigth forward.
You pass: 

- the object(a list), 
- the function to be applied to all elements of that list, 
- and any additional arguments to that function


Exercises:
```{r, eval=FALSE, include=FALSE}

require(purrr)
# using the `reduce()` function

# 1. Store all objects you want to combine in a list object
tables <- list(table1, table2, table3, table4)
reduce(tables, left_join, by="name")

# This resembles the following pipe
table1 %>% 
        left_join(table2, by="name") %>% 
        left_join(table3, by="name") %>% 
        left_join(table4, by="name")
```

##### Example Exercises to the purr functions:
```{r}
# 
# 1. Place supergroups, more_bands, and more_artists into a list
list(supergroups, more_bands, more_artists) %>% 
  
        # Use reduce to join together the contents of the list
  reduce(left_join)


# It works with all dplyr functions 
list(more_artists, more_bands, supergroups) %>% 
  # Return rows of more_artists in all three datasets
  reduce(semi_join, by=c("first", "last"))
```

All these joins can be computed using only the base-R Package using the `merge()` function:
```{r, eval=FALSE, include=FALSE}
# merge is a join super function, enabling all types of joins by specifying the arguments in the function call.

# left_join(names, plays, by="name")
merge(names, plays, by="name", all.x=TRUE, all.y=FALSE)

# right_join(names, plays, by="name")
merge(names, plays, by="name", all.x=FALSE, all.y=TRUE)

# inner_join(names, plays, by="name")
merge(names, plays, by="name", all.x=FALSE, all.y=FALSE)

# full_join(names, plays, by="name")
merge(names, plays, by="name", all=TRUE)
```



