---
title: "R Notebook"
output: html_notebook
---
```{r}
require(dplyr)
```

Setting up a dummy dataset
```{r}
IQ <- as.data.frame(matrix(rnorm(100, 100,15), 10,10))
names(IQ) <- sprintf("Test%s", 1:10)

# Splitting the table
table1 <- IQ[seq(1,dim(IQ)[1],2),]
table2 <- IQ[seq(2,dim(IQ)[1],2),]

# Checking the resulting dimensions
dim(table1)
dim(table2)
table1
```

## Basic Datasetcombination : Data-Binding

When we want to combine parts containing the exact same length of observations on the same entries, then we can just combine them.
This is done with `binds`
```{r}
# The base-package functions are ...
rbind()
cbind()

# Dplyr offers vectorized and advanced functions to do the same

# to datasets with exact the same number of columns into a single dataset
bind_rows(table1, table2)
bind_cols(table1, table2)
```

What are reasons to use the dplyr functions instead of the base-r package ones?

The dplyr functions are implemented to be:

- faster, as they directly implement in C++ Function calls
- the always `return a tibble`
- can `handle lists of data frames` as input. so no need for a loop
- offer the .id attribute

We can use the .id attribute to store the information of the originating dataframes as a value in a new formed colomn. 
This is done like this...

```{r}
# Again joining the same tables as before, but now indicating that they resulted from the testing of two different groups.
# One is now the control, one is the test group...

table_rec <- bind_rows(Controll = table1, Test = table2, .id = "Group")
bind_rows(Controll = table1, Test = table2, .id = "Group")

# Filter now to select subsets
table_rec %>% filter(Group=="Controll") %>% select(-Group) %>% sapply(sd)
```

## Building better data.frames in R using `data_frame()` and `as_data_frame()`

The base-r functions:

- `data.frame()`:       
-`as.data.frame()`:     

data.frame might change columns names, data.types or 

Provide possibilities to transform lists of vectors into a data.frame type to be use
```{r}
# Using data_frame() we pass the columns as concatenated arguments
data_frame(
        Beatles = c("John", "Paul", "George","Ringo"),
        Stones = c("Mick", "Keith", "Charlie", "Ronnie"),
        Zeppelins = c("Robert", "Jimmy", "John Paul", "John")
)
```
This function will NEVER:

- Change the data type of oyur vectors(e.g. strings to factors)
- Add row names
- Change column names
- Recycle vectors greater than length one

Therefore, we have to specify all arguments by hand. 

It does:

- Evaluate arguments in order lazily, therefore letting us refer to columns `specified prior in the same call`
```{r}
data_frame(
        numbers = 1:5,
        squares = numbers ^ 2
)
```

- Outputs a tibble (an enhanced data.frame)


### as_data_frame() converts a list of columns into a dataframe

It works as data_frame()
```{r}
# setting up a list of lists
result <- list(10)
for(i in 1:10){ result[[i]] <- 1:100}
names(result) <- sprintf("Variable %s", 1:10)

# Converting lists into data.frame
as_data_frame(result)
```


When combining a set of columns in a list we can easily use the data_frame() function. 

But, when we have to handle the combination of multiple data.frames bound in a list, `data_frame()` provides no solution. 
Here we can use the `dplyr` function `bind_rows()`. 

As this function is vectorized, it can handle lists automatically, saving us from writing out loops over the list.

Example:
```{r}

# Lets take this
result <- list(10)
for(i in 1:100){ result[[i]] <- as.data.frame(matrix(rnbinom(100,2500,0.5), 10,10)) }

result <- lapply(result, function(x){ colnames(x) <- sprintf("Test %s", 1:10);x})

# If we now try to pass this list of dataframes to the as_data_frame() function, we will get an error
tryCatch(as_data_frame(result),
         error=function(e){
                 message("Error occured trying to connect the dataframes: ",e)
         })

# Trying to do the same with dplyr
result <- bind_rows(result) 
sapply(result, mean)
```

### Solutions for coercing datatypes problems

when dataframes are joined, the datatypes sometimes have to be automatically joined together.
This process is made easy in dplyr, as it will try to help automatically overcome the most notorious problems.


```{r}

a <- 1:5
as.factor(a, levels(c("A", "B", "C", "D", "E")))


```

